The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-09-19 sections/climate.tex
Implement corrections from Irina
Mostly as to-do notes for now

Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalysis and one regional reanalysis.

@@

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer. Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. From an ice sheet mass balance point of view, this contrast becomes even stronger by the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

There are two problems inherent to the use of WorldClim data in our study area. Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. Secondly, WorldClim lacks of data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

>>>>

Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.

@@

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer. Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. From an ice sheet mass balance point of view, this contrast becomes even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

There are two problems inherent to the use of WorldClim data in our study area. Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. Secondly, WorldClim lacks data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

### 2013-09-19 sections/concl.tex
Implement corrections from Qiong
Implement corrections from Irina
Mostly as to-do notes for now

\emph{Coming soon...}

>>>>

Our study shows a strong dependency of the modelled ice sheet on the choice of climate forcing data. When using input climatologies from climate reanalyses, spatial resolution of the GCM appears critical for providing the ice sheet model with an accurate precipitation field. This confirms results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.

For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}. The latter dataset is preferable in our case due to the lack of WorldClim data offshore. NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.

One must keep in mind, however, that these results are biased by our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets). At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation. At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.
\irina{I have this feeling that the manuscript contains a number of repetitive ideas and arguments. I will try to track them throughout the text.}

### 2013-09-19 sections/discuss.tex
Implement corrections from Irina
Mostly as to-do notes for now

Our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean. These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}). This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution. In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).

>>>>

Our simulations produce anomalous\irina{Any evidence that there was no ice cover there? Refer to the source.} ice cover on parts of the continental shelf in the Arctic Ocean. These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}). This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution. In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).

### 2013-09-19 sections/model.tex
Implement corrections from Irina
Mostly as to-do notes for now

We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}. Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response. Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except of the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is further computed in 11 regularly spaced layers in bedrock. A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.

@@

Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature. This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections applies to precipitation.

>>>>

We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}. Given basal topography\irina{geothermal heat flux and sea level} and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response. Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except from the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice, and temperature is further computed in 11 regularly spaced layers in bedrock. A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.

@@

Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature. This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections apply to precipitation.

### 2013-09-19 sections/results.tex
Implement corrections from Irina
Mostly as to-do notes for now

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}. Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig. \ref{fig:ivolarea}-\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 \ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{\degree C} offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent is primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.

To contrast the six different input climatologies, we plot simulations run with the a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

>>>>

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig. \ref{fig:ivolarea}-\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{\degree C} offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

To contrast the six different input climatologies, we plot simulations run with the 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.
